### 2021-12-08 manuscript.tex
Simplify article table [ci skip]
This build is based on
https://github.com/lubianat/quali_phd/commit/8e4db1e93b09d5722f9c3655e461f238c65e0391.

This commit was created by the following CI build and job:
https://github.com/lubianat/quali_phd/commit/8e4db1e93b09d5722f9c3655e461f238c65e0391/checks
https://github.com/lubianat/quali_phd/actions/runs/1555928912

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/b7d3fd0aa464eb239683d617aa20c9cb1a738276/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/b7d3fd0aa464eb239683d617aa20c9cb1a738276}{lubianat/quali\_phd@b7d3fd0}
on December 8, 2021.

@@

\hypertarget{status-of-cell-type-info-on-wikidata}{%
\subsubsection{Status of cell type info on Wikidata}\label{status-of-cell-type-info-on-wikidata}}

Status of cell type information on Wikidat was acessed via SPARQL queries combined with processing in python and is available at \url{https://colab.research.google.com/drive/1GvQXOs51_U8icdGMtKXMeLOXKM8pXWet\#scrollTo=szvBWI9zr_AA}

@@

Refining the different concepts around the notion of ``cell type'' is important, but will require decades before a reasonable consensus.
Here we adopt a liberal view of cell type, defining, for our purposes, a cell type as ``any class of cells described by a domain expert''.
By ``class'' we mean an abstract entity in the sense intended by the multilevel theory of conceptual modelling (referred as ``\emph{types}'' by Carvalho et al.~) {[}\protect\hyperlink{ref-eP47FILn}{114}{]}

Figure \ref{fig:multilevel} represents the adaptation of Multi Level Theory that is assumed for the modelling done throughout this work.
The notion is time-agnostic (i.e.~it is not concerned if their objects \emph{exist now} or just \emph{existed in the past}) and organizes the theoretical world into 3 kingdoms: \emph{individuals}, \emph{classes} and \emph{metaclasses}.

@@

\begin{itemize}
\item
  Justifying the selection: MLT and basic formal ontology
\item
  Justifying the selection: MLT and wikidata
\end{itemize}

@@

The process of making the Human Cell Atlas more useful via Wikidata also includes the connection of related databases.
PanglaoDB {[}\protect\hyperlink{ref-M0lqRYJb}{115}{]} {[}\protect\hyperlink{ref-T3PlZ3Vh}{116}{]} is a publically-available database that contains data and metadata on hundreds of single-cell RNA sequencing experiments.
It provides extensive information on cell types, genes, and tissues and cell type markers, obtained both via automatic and manual methods.
It also displays a rich web user interface for easy data acquisition, including database dumps for bulk downloads.

As of 8 December June 2021, the article describing PanglaoDB had been cited 230 times.
Despite its use by the community, the database is on a 3-star category for Linked Open Data {[}\protect\hyperlink{ref-OGXYtGZ8}{117}{]} as it does not use the open semantic standards from W3C (RDF and SPARQL) needed for a 4-star rank, neither the links to external data via standard identifiers that make datasets 5-star.
Improving the data format toward W3C's gold standards is a valuable step in making biological knowledge FAIR (Findable, Accessible, Interoperable, and Reusable).
Thus, we aimed to provide a case study of making the core information of PandlaoDB available in a 5-star Linked Open Data Format while improving the modeling of the necessary concepts on Wikidata.

As of August 2020, Wikidata had 264 items being categorized as a ``cell type'', considerably less than in specialized cell catalogs, which count over two thousand cell types {[}\protect\hyperlink{ref-agnqfdk6}{45},\protect\hyperlink{ref-4AEy2xhQ}{118}{]}.
Strikingly, there were also 23 items categorized as ``instances of cell (Q7868)''. This classification is imprecise, as an instance of cell would be an individual named cell from a single named individual, an example of conceptual disarray that often occurs on Wikidata. {[}\protect\hyperlink{ref-imSZTi52}{119}{]}

@@

We diligently fixed and improved information on cell types on Wikidata.
As of 8 December 2021, the Wikidata database contains 2834 instances of ``cell type'' (see current status at \url{https://w.wiki/b2t}) and 0 instances of ``cell'' (\url{https://w.wiki/4XAg}) highlighting the improvements in both quantity and quality.

@@

Now that we re-formatted the markers on PanglaoDB as Linked Open Data, we can make queries that were not possible before, includingfederated queries with other biological databases, such as Uniprot {[}\protect\hyperlink{ref-HUJ9NzAb}{120}{]}
and Wikipathways {[}\protect\hyperlink{ref-AgOOCR24}{121}{]}.
Due to previous similar reconciliation projects, Wikidata already contains information about genes, including their relations to Gene Ontology (GO) terms.

@@

This work exemplifies the power of releasing Linked Open Data via Wikidata, and provides the biomedical community with the first semantically accessible, 5-star LOD dataset of cell markers, easily reachable from Wikidata's SPARQL Query Service (\url{https://query.wikidata.org/}).
It is not first case study of biomedical data integration to Wikidata (see {[}\protect\hyperlink{ref-mPoPwN77}{103}{]} for example.
Nevertheless, the differences among the articles in style and scope contribute to a richer ecosystem for possible contributor.
{]})
The work also paves the way for Wikidata reconciling of other databases for cell-type markers, such as CellMarker {[}\protect\hyperlink{ref-chGii6yw}{43}{]}, labome {[}\protect\hyperlink{ref-rhRRCtlA}{122}{]}, CellFinder {[}\protect\hyperlink{ref-4AEy2xhQ}{118}{]} and SHOGoiN/CELLPEDIA {[}\protect\hyperlink{ref-6uWWsiSq}{123}/{]}) (if proper authorization are given by the owners).
The approach we took here can in essence be applied to any knowledge set of public interest, providing a low-cost and low-barrier platform for sharing biocurated knowledge in gold standard format.

@@

In the humanities, there is a well-established practice of annotations of readings.
The annotation skills are part of common academic training in the humanities {[}\protect\hyperlink{ref-rPKBwmYh}{124}/{]}{[}\protect\hyperlink{ref-PKhuVRW8}{125}\_da26C-QW5qiS7uZ{]}.
An influential work in presenting methods for academic reading in the humanities is Umberto Eco's book ``How to Write a Thesis'' {[}\protect\hyperlink{ref-1HBVPtZGp}{126}{]}, which outlines not only \emph{how} to annotate the literature that basis an academic thesis, but also \emph{why} to do so.
The book, written originally in 1977, is still influential today, but its theoretical scope (roughly the humanities) and its date, preceding the digital era, limits the extent in which it applies to the biomedical sciences.

Notably, the need of an organized reading system for biocuration studies stems from a difference in methodology.
In humanities, the main (if not sole) research material is the written text, the books and articles from which research stems. {[}\protect\hyperlink{ref-PKhuVRW8}{125}\_da26C-QW5qiS7uZ{]}.
In the biomedical sciences, including a large part of bioinformatics, the object of study is the natural world, observed via experimentation.
Thus, naturally, scientific training focuses on the theoretical and practical basis of experimentation and data analysis.
With the bloom of scientific articles, however, the scientific literature (and accompaning public datasets) provide already a strong material for the sculpting of scientific projects.
Thus, the development of a methodology for academic reading, tailored to the digital environment, presents itself as a need.

@@

The \texttt{docs/} directory contains the live dashboard from the readings, which will be discussed in the following sessions.
The \texttt{downloads/} directory hosts the pdfs of the articles read with the system.
These are not commited to the repository, and are only stored locally.
The \texttt{notes/} directory contains markdown files, one for each article read.
The \texttt{src/} directory contains the python code with the mechanics of the system.
They contain helper functions for the command line commands discussed below:
- \texttt{wread} which receives a Wikidata QID for an article and outputs (1) a notes document, (2) a pdf for the paper obtained from Unpaywall {[}\protect\hyperlink{ref-15luL9zZC}{127}/{]} and (3) an updated version of the dashboard html files in the \texttt{docs/} directory.
- \texttt{pop}, which ``pops'' an article from \texttt{toread.md} and runs \texttt{wread} for it
- \texttt{wadd}, which takes an URL for an Wikidata SPARQL query and adds new QIDs to \texttt{toread.md}
- \texttt{wadd\_all}, which parses \texttt{config.yaml} for recurrent SPARQL queries and runs \texttt{wadd} for each
- \texttt{wlog}, which adds, commits and pushes recent readings and dashboard updates to GitHub

All the structures described so far are commonly shared by any user of Wikidata Bib.
To personalize the use of the system, the user edits three plain text files.
\texttt{toread.md} hosts a plain text QIDs of the articles that will be read.
These can be added either manually, or via wadd.
While the \texttt{pop} command only sees QIDs, articles titles or other identifiers can be added to \texttt{toread.md} temporarily without breaking the system.
\texttt{index.md} hosts a numbered list of topics of interest.
This file plays the role of Umberto Eco's work plan, with the topics of interest for the academic. {[}\protect\hyperlink{ref-1HBVPtZGp}{126}{]}
These are used to tag articles for retrieval in a later step.
\texttt{config.yaml} contains shortcuts for different reading lists.
This is better explained by example.
In my \texttt{toread.md} file there are two reading lists, one following a \texttt{\#\ Cell\ types} header, and another following a \texttt{\#\ Biocuration} header.
My \texttt{config.yaml} contains the following snippet:

@@

The Wikidata bib framework is coupled with a discipline of daily reading.
This is inspired by Robert Cecil Martin's description of Test Driven Development in the book ``Clean Code'', which includes not only a technical description, but a \emph{school of thought} of how software development can be approached. {[}\protect\hyperlink{ref-13HqB23xH}{128}{]}
Every day, I read one article of each list, using the notetaking station displayed in Figure \ref{fig:notetaking}.
The constancy of reading allows steady coverage of the relevant literature.
While it has worked for this research project, however, it is not required for use of the Wikidata Bib system.

@@

The discipline also includes, whenever possible, an improvement of the metadata about the article on Wikidata.
In \ref{fig:notetaking} B are shown the links included in the dashboard.
A link to a Scholia {[}\protect\hyperlink{ref-hxzL9pmm}{100}{]} profile allows identification of related articles from a series of pre-made SPARQL queries probing bibliography data on Wikidata.
While Scholia provides an overview of a given article, it does not allow direct curation of the metadata.
For that, two links are provided, one to Wikidata and one to Author Disambiguator {[}\protect\hyperlink{ref-1A9RvszKC}{129}{]}.
By acessing the Wikidata page for the entity, one can add new triples, for example curating authors and topics of the article, which are then used by Scholia and by Wikidata Bib's dashboard.
Author Disambiguator is a wrapper of an Wikimedia API which facilitates the process of disambiguating author names to unique identifiers on Wikidata, thus feeding the publict knowledge graph of publication and authors.\\
Finally, a link to the article's DOI or full text URL is provided, and serves as a fallback when the automatic download fails.
Of note, while the metadata curation has a technical benefit to Wikidata and the dashboard, it also plays a theoretical role.
By curating metadata on authors, the user of Wikidata Bib can better understand the people they read, and expand their metascientific perspective on their domain of interest.

@@

To organize the code for rendering the dashboard, we created a python package, wbib, and deposited it in PyPi, making it available via \texttt{pip}. {[}\protect\hyperlink{ref-6chnW6cc}{130}/{]}.
The package implements the logic for rendering complex Wikidata-based academic dashboards and is available in GitHub at https://github.com/lubianat/wbib.
It allows the user to build dashboards based on Wikidata records of information such as gender of authors, the region of author\texttt{s\ intitutions,\ topics\ of\ articles\ and\ similar\ metascientific\ information.\ \ The\ dashboard\ is\ composed\ of\ SPARQL\ queries\ written\ for\ the\ Wikidata\ Query\ Service\ {[}@url:https://query.wikidata.org{]}\ \ It\ also\ allows\ users\ to\ feed\ an\ arbitrary\ list\ of\ articles\ and\ obtain\ a\ custom\ dashboard.\ \ Wikidata\ Bib\ obtains\ the\ html\ dashboards\ after\ feeding\ wbib\ the\ lists\ of\ articles\ read\ in\ total\ (}index.html\texttt{)\ or\ in\ pre-determined\ time\ spans\ (}last\_day.html\texttt{,}past\_week.html\texttt{and}past\_month.html` )

@@

The dashboard includes not only a basic list of read articles, but also statistics on most read authors and most read venues.
It also displays an interactive map of the instituions of articles read, permitting a glance on geographic biases in activities.
An example of queries is show in \ref{fig:dashboard}.
As the queries are rendered live, they evolve in quality with the growth of Wikidata.
Finally, the clean 5-star-open data format enables users to adapt the queries to include different aspects of Wikidata.
For example, table \ref{tbl:articles_read_hca} showcases 10 articles that (1) I have read in the past year and (2) were authored by a speaker of the 1st Human Cell Atlas Latin America Single Cell RNA-seqData Analysis Workshop {[}\protect\hyperlink{ref-1hag8XE6}{131}/{]}.
One practical application that the dashboard enables, thus, is to identify people in an event, institution or location that the user has read before, therefore catalysing the possibility of colaborations.
Anedoctally, this strategy was tested successfully at Biohackathon Europe 2021 {[}\protect\hyperlink{ref-kHL3NVxk}{132}/{]}, where I used the system both to identify possible collaborators and as a conversation starter.

\begin{tablenos:no-prefix-table-caption}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2286}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0974}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.6740}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
workLabel
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
authors\_in\_the\_conference
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
DOI
\end{minipage} \\
\midrule
\endhead
A promoter-level mammalian expression atlas & Jay W Shin & https://doi.org/10.1038/NATURE13182 \\
Single-cell RNA-seq reveals new types of human blood dendritic cells, monocytes, and progenitors. & Muzlifah Haniffa & https://doi.org/10.1126/SCIENCE.AAH4573 \\
The Human Cell Atlas. & Musa Mhlanga, Jay W Shin, Muzlifah Haniffa, Menna R Clatworthy, Dana Pe'er & https://doi.org/10.7554/ELIFE.27041 \\
The Human Cell Atlas: Technical approaches and challenges. & Jay W Shin & https://doi.org/10.1093/BFGP/ELX029 \\
Innate Immune Landscape in Early Lung Adenocarcinoma by Paired Single-Cell Analyses. & Dana Pe'er & https://doi.org/10.1016/J.CELL.2017.04.014 \\
Single cell RNA sequencing of human liver reveals distinct intrahepatic macrophage populations & Sonya A MacParland & https://doi.org/10.1038/S41467-018-06318-7 \\
Single-cell reconstruction of the early maternal--fetal interface in humans & Muzlifah Haniffa & https://doi.org/10.1038/S41586-018-0698-6 \\
Distinct microbial and immune niches of the human colon & Rasa Elmentaite, Menna R Clatworthy & https://doi.org/10.1038/S41590-020-0602-Z \\
A cell atlas of human thymic development defines T cell repertoire formation & Muzlifah Haniffa, Menna R Clatworthy & https://doi.org/10.1126/SCIENCE.AAY3224 \\
Decoding human fetal liver haematopoiesis & Muzlifah Haniffa & https://doi.org/10.1038/S41586-019-1652-Y Table: Articles read by Tiago Lubiana before 8 December 2021 in which an author was a speaker at HCA Latin America \\
\{\#tbl:articles\_read\_hca\} & & \\
\bottomrule
\end{longtable}

\end{tablenos:no-prefix-table-caption}

>>>>

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/8e4db1e93b09d5722f9c3655e461f238c65e0391/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/8e4db1e93b09d5722f9c3655e461f238c65e0391}{lubianat/quali\_phd@8e4db1e}
on December 8, 2021.

@@

\hypertarget{status-of-cell-type-info-on-wikidata-and-the-cell-ontology}{%
\subsubsection{Status of cell type info on Wikidata and the Cell Ontology}\label{status-of-cell-type-info-on-wikidata-and-the-cell-ontology}}

Status of cell type information on Wikidat was acessed via SPARQL queries combined with processing in python and is available at \url{https://colab.research.google.com/drive/1GvQXOs51_U8icdGMtKXMeLOXKM8pXWet\#scrollTo=szvBWI9zr_AA}.

Counts of cell classes in the Cell Ontology were performed via regex matching on Cell Ontology releases following the code available at https://github.com/lubianat/cell\_ontology\_count.

@@

Refining the different concepts around the notion of ``cell type'' is important, but will require decades before a reasonable consensus. Here we adopt a liberal view of cell type, defining, for our purposes, a cell type as any class of cells described by a domain expert with evidence of reality of its instances.
The requirement of evidence of existence in reality is based on the Principle of instantiation of ontological realism {[}\protect\hyperlink{ref-5ZFRyTy5}{114}{]}. Barry Smith and Werner Ceuster states that ``A term should be included in a reference ontology only if there is experimental evidence that instances to which that term refers exist in reality.(`Exists' here should be understood in a tenseless sense in order to accommodate, for example, universals pertaining to extinct species as well as universals such as swarm or hurricane which are instantiated only intermittently.)''
Thus, in this work one minimum requirement for a cell type to be catalogued is a public description by a researcher of the class, with evidence for existence of instances of the class in reality.

By ``class'' we mean an abstract entity in the sense intended by the multilevel theory (MLT) of conceptual modelling (referred as ``\emph{types}'' by Carvalho et al.~) {[}\protect\hyperlink{ref-eP47FILn}{115}{]}
Figure \ref{fig:multilevel} displays a simplified version of MLT adopted throughout this project.
In this framework, real-world entities are materializations of \emph{individuals}.
\emph{Individuals} are theoretical constructs which are (1) thought to exist or have existed, as per the principle of instantiation, and (2) refer to only one (01) material entity at any point in time.
For example, Wikidata has entries for people, e.g.~``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' which are considered \emph{individuals} by Multi Level Theory.
Other examples of \emph{individuals} include ``Albert Einstein's brain (\href{https://www.wikidata.org/wiki/Q2464312}{Q2464312})'' and the ``Christ the Redeemer statue (\href{https://www.wikidata.org/wiki/Q79961}{Q79961})''.

@@

In MLT, \emph{individuals} are instances of some \emph{classes}.
For example, both ``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' could be represented as instances of the class ``\emph{Homo sapiens} (\href{https://www.wikidata.org/wiki/Q15978631}{Q15978631})'' on Wikidata.
``\emph{Homo sapiens} (Q15978631)'' is only one of the classes that those individuals belong to.
Another one is ``animal (\href{https://www.wikidata.org/wiki/Q729}{Q729})''.
As all instances of ``\emph{Homo sapiens} (Q15978631)'' are also instances of ``animal (Q729)'', ``\emph{Homo sapiens} (Q15978631)'' is a subclass of ``animal (Q729)''.
It is possible to continue the hierarchy of subclasses, as ``animal (Q729)'' is a subclass of ``organism (Q7239)'', until the root case, which in the case of wikidata is the class ``entity (\href{https://www.wikidata.org/wiki/Q35120}{Q35120})''

Classes, however, can themselves behave as individuals.
For example, both ``\emph{Homo sapiens} (Q15978631)'' and ``animal (Q729)'' are instances of ``taxon (\href{https://www.wikidata.org/wiki/Q16521}{Q16521})''.
``Taxon (Q16521)'', thus, is a \emph{metaclass}, or, more precisely, a \emph{1st-order metaclass}.
Other examples of metaclasses are ``\emph{species} (\href{https://www.wikidata.org/wiki/Q7432}{Q7432})'' and ``phylum (\href{https://www.wikidata.org/wiki/Q38348}{Q38348})''.
These, in turn, are instances of ``taxonomic rank (\href{https://www.wikidata.org/wiki/Q427626}{Q427626})'', a \emph{2nd-order metaclass}.

In the Figure \ref{fig:multilevel} B there is a proposal of this version of MLT for cell types.
As individual cells are rarely named, for the sake of example, we can consider the ``zygote of Mahatman Gandhi'' as an \emph{individual} in the theoretical system, an instance of the class ``zygote (\href{https://www.wikidata.org/wiki/Q170145}{Q170145})'', which is itself an instance of the metaclass ``cell type (\href{https://www.wikidata.org/wiki/Q189118}{Q189118})''
A more concrete example stems from RNA-sequencing datasets with barcodes for each single cell in a particular sample.
Each barcode can be thought as an identifier for an \emph{individual}.
Thus, labeling single-cells is a process of identification, where each \emph{individual} is connected to a \emph{class} of interest.

For the practical purpose adopted here, we avoid the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{111}{]} for an example). We also consider only the cell as it was observed in an experiment, not necessarily the future conditions of any cell (i.e., the ``cell fate''). {[}\protect\hyperlink{ref-OjxQDZpx}{27}{]}\\
Even though such a distinction is an important topic for theoretical research, it is outside the initial scope of this work.

Another logical consequence of the definition is that the concept of subtype becomes redundant with the concept of cell type.
The notion of subtype, then, only makes sense when discussing classes with different degrees of universality.
Thus, claims to discover new cell ``subtypes'' or ``types'' differ only stylistically and can be considered indistinguishable from the perspective of research synthesis.

We also note that we made a judgment call to use the term ``cell type'' to emphasize the focus on types as classes (or ``kinds'') in contrast to real-world objects.
The term ``cell class'' is also used in the literature and is a suitable synonym for our notion of cell type.
We opted to frame our work around the term ``cell type'' due to its historical usage and familiarity for the life sciences community.
Other related thems a ``cell set,'' ``cell population,'' and ``cell cluster,'' can also reminisce of a specific, countable group of cells, frequently from the same experiment.
The term ``cell identity'' has also been suggested for avoiding the cell type/cell state dilemma {[}\protect\hyperlink{ref-WKbly37M}{38}{]}, but we avoid it to emphasize a nominalistic perspective (in the Popperian sense{[}\protect\hyperlink{ref-1C6LI68h6}{116}{]}).
In doing so, we reinforce the intent on represent on what cell types are \emph{reported to exist}, instead of trying to state bluntly which cell types \emph{exist} or, even worse, are \emph{essential} of human beings.

The employment of MLT as described before, and the notion of species-specific cell types are fundamental for the next chapters of this work.
In the chapter about the PanglaoDB integration, we describe how we used the notion of species-specific types to add marker information to Wikidata, and how we cleaned up conceptual disarrays that broke MLT.

@@

The process of making the Human Cell Atlas more useful via Wikidata also includes the connection of related databases.
PanglaoDB {[}\protect\hyperlink{ref-M0lqRYJb}{117}{]} {[}\protect\hyperlink{ref-T3PlZ3Vh}{118}{]} is a publically-available database that contains data and metadata on hundreds of single-cell RNA sequencing experiments.
It provides extensive information on cell types, genes, and tissues and cell type markers, obtained both via automatic and manual methods.
It also displays a rich web user interface for easy data acquisition, including database dumps for bulk downloads.

As of 8 December June 2021, the article describing PanglaoDB had been cited 230 times.
Despite its use by the community, the database is on a 3-star category for Linked Open Data {[}\protect\hyperlink{ref-OGXYtGZ8}{119}{]} as it does not use the open semantic standards from W3C (RDF and SPARQL) needed for a 4-star rank, neither the links to external data via standard identifiers that make datasets 5-star.
Improving the data format toward W3C's gold standards is a valuable step in making biological knowledge FAIR (Findable, Accessible, Interoperable, and Reusable).
Thus, we aimed to provide a case study of making the core information of PandlaoDB available in a 5-star Linked Open Data Format while improving the modeling of the necessary concepts on Wikidata.

As of August 2020, Wikidata had 264 items being categorized as a ``cell type'', considerably less than in specialized cell catalogs, which count over two thousand cell types {[}\protect\hyperlink{ref-agnqfdk6}{45},\protect\hyperlink{ref-4AEy2xhQ}{120}{]}.
Strikingly, there were also 23 items categorized as ``instances of cell (Q7868)''. This classification is imprecise, as an instance of cell would be an individual named cell from a single named individual, an example of conceptual disarray that often occurs on Wikidata. {[}\protect\hyperlink{ref-imSZTi52}{121}{]}

@@

We diligently fixed and improved information on cell types on Wikidata.
As of 8 December 2021, the Wikidata database contains 2834 instances of ``cell type'' (see current status at \url{https://w.wiki/b2t}) and 0 instances of ``cell'' (\url{https://w.wiki/4XAg}) highlighting the improvements in both quantity and quality.
This increase stems both from the PanglaoDB initiative (around 430 new types) and from the Wikidata Bib curation, described in a later chapter.

@@

Now that we re-formatted the markers on PanglaoDB as Linked Open Data, we can make queries that were not possible before, includingfederated queries with other biological databases, such as Uniprot {[}\protect\hyperlink{ref-HUJ9NzAb}{122}{]}
and Wikipathways {[}\protect\hyperlink{ref-AgOOCR24}{123}{]}.
Due to previous similar reconciliation projects, Wikidata already contains information about genes, including their relations to Gene Ontology (GO) terms.

@@

This work exemplifies the power of releasing Linked Open Data via Wikidata, and provides the biomedical community with the first semantically accessible, 5-star LOD dataset of cell markers, easily reachable from Wikidata's SPARQL Query Service (\url{https://query.wikidata.org/}).
It is not first case study of biomedical data integration to Wikidata (see {[}\protect\hyperlink{ref-mPoPwN77}{103}{]} for example.
Nevertheless, the differences among the articles in style and scope contribute to a richer ecosystem for possible contributor.
{]})
The work also paves the way for Wikidata reconciling of other databases for cell-type markers, such as CellMarker {[}\protect\hyperlink{ref-chGii6yw}{43}{]}, labome {[}\protect\hyperlink{ref-rhRRCtlA}{124}{]}, CellFinder {[}\protect\hyperlink{ref-4AEy2xhQ}{120}{]} and SHOGoiN/CELLPEDIA {[}\protect\hyperlink{ref-6uWWsiSq}{125}/{]}) (if proper authorization are given by the owners).
The approach we took here can in essence be applied to any knowledge set of public interest, providing a low-cost and low-barrier platform for sharing biocurated knowledge in gold standard format.

@@

In the humanities, there is a well-established practice of annotations of readings.
The annotation skills are part of common academic training in the humanities {[}\protect\hyperlink{ref-rPKBwmYh}{126}/{]}{[}\protect\hyperlink{ref-PKhuVRW8}{127}\_da26C-QW5qiS7uZ{]}.
An influential work in presenting methods for academic reading in the humanities is Umberto Eco's book ``How to Write a Thesis'' {[}\protect\hyperlink{ref-1HBVPtZGp}{128}{]}, which outlines not only \emph{how} to annotate the literature that basis an academic thesis, but also \emph{why} to do so.
The book, written originally in 1977, is still influential today, but its theoretical scope (roughly the humanities) and its date, preceding the digital era, limits the extent in which it applies to the biomedical sciences.

Notably, the need of an organized reading system for biocuration studies stems from a difference in methodology.
In humanities, the main (if not sole) research material is the written text, the books and articles from which research stems. {[}\protect\hyperlink{ref-PKhuVRW8}{127}\_da26C-QW5qiS7uZ{]}.
In the biomedical sciences, including a large part of bioinformatics, the object of study is the natural world, observed via experimentation.
Thus, naturally, scientific training focuses on the theoretical and practical basis of experimentation and data analysis.
With the bloom of scientific articles, however, the scientific literature (and accompaning public datasets) provide already a strong material for the sculpting of scientific projects.
Thus, the development of a methodology for academic reading, tailored to the digital environment, presents itself as a need.

@@

The \texttt{docs/} directory contains the live dashboard from the readings, which will be discussed in the following sessions.
The \texttt{downloads/} directory hosts the pdfs of the articles read with the system.
These are not commited to the repository, and are only stored locally.
The \texttt{notes/} directory contains markdown files, one for each article read.
The \texttt{src/} directory contains the python code with the mechanics of the system.
They contain helper functions for the command line commands discussed below:
- \texttt{wread} which receives a Wikidata QID for an article and outputs (1) a notes document, (2) a pdf for the paper obtained from Unpaywall {[}\protect\hyperlink{ref-15luL9zZC}{129}/{]} and (3) an updated version of the dashboard html files in the \texttt{docs/} directory.
- \texttt{pop}, which ``pops'' an article from \texttt{toread.md} and runs \texttt{wread} for it
- \texttt{wadd}, which takes an URL for an Wikidata SPARQL query and adds new QIDs to \texttt{toread.md}
- \texttt{wadd\_all}, which parses \texttt{config.yaml} for recurrent SPARQL queries and runs \texttt{wadd} for each
- \texttt{wlog}, which adds, commits and pushes recent readings and dashboard updates to GitHub

All the structures described so far are commonly shared by any user of Wikidata Bib.
To personalize the use of the system, the user edits three plain text files.
\texttt{toread.md} hosts a plain text QIDs of the articles that will be read.
These can be added either manually, or via wadd.
While the \texttt{pop} command only sees QIDs, articles titles or other identifiers can be added to \texttt{toread.md} temporarily without breaking the system.
\texttt{index.md} hosts a numbered list of topics of interest.
This file plays the role of Umberto Eco's work plan, with the topics of interest for the academic. {[}\protect\hyperlink{ref-1HBVPtZGp}{128}{]}
These are used to tag articles for retrieval in a later step.
\texttt{config.yaml} contains shortcuts for different reading lists.
This is better explained by example.
In my \texttt{toread.md} file there are two reading lists, one following a \texttt{\#\ Cell\ types} header, and another following a \texttt{\#\ Biocuration} header.
My \texttt{config.yaml} contains the following snippet:

@@

The Wikidata bib framework is coupled with a discipline of daily reading.
This is inspired by Robert Cecil Martin's description of Test Driven Development in the book ``Clean Code'', which includes not only a technical description, but a \emph{school of thought} of how software development can be approached. {[}\protect\hyperlink{ref-13HqB23xH}{130}{]}
Every day, I read one article of each list, using the notetaking station displayed in Figure \ref{fig:notetaking}.
The constancy of reading allows steady coverage of the relevant literature.
While it has worked for this research project, however, it is not required for use of the Wikidata Bib system.

@@

The discipline also includes, whenever possible, an improvement of the metadata about the article on Wikidata.
In \ref{fig:notetaking} B are shown the links included in the dashboard.
A link to a Scholia {[}\protect\hyperlink{ref-hxzL9pmm}{100}{]} profile allows identification of related articles from a series of pre-made SPARQL queries probing bibliography data on Wikidata.
While Scholia provides an overview of a given article, it does not allow direct curation of the metadata.
For that, two links are provided, one to Wikidata and one to Author Disambiguator {[}\protect\hyperlink{ref-1A9RvszKC}{131}{]}.
By acessing the Wikidata page for the entity, one can add new triples, for example curating authors and topics of the article, which are then used by Scholia and by Wikidata Bib's dashboard.
Author Disambiguator is a wrapper of an Wikimedia API which facilitates the process of disambiguating author names to unique identifiers on Wikidata, thus feeding the publict knowledge graph of publication and authors.\\
Finally, a link to the article's DOI or full text URL is provided, and serves as a fallback when the automatic download fails.
Of note, while the metadata curation has a technical benefit to Wikidata and the dashboard, it also plays a theoretical role.
By curating metadata on authors, the user of Wikidata Bib can better understand the people they read, and expand their metascientific perspective on their domain of interest.

@@

To organize the code for rendering the dashboard, we created a python package, wbib, and deposited it in PyPi, making it available via \texttt{pip}. {[}\protect\hyperlink{ref-6chnW6cc}{132}/{]}.
The package implements the logic for rendering complex Wikidata-based academic dashboards and is available in GitHub at https://github.com/lubianat/wbib.
It allows the user to build dashboards based on Wikidata records of information such as gender of authors, the region of author\texttt{s\ intitutions,\ topics\ of\ articles\ and\ similar\ metascientific\ information.\ \ The\ dashboard\ is\ composed\ of\ SPARQL\ queries\ written\ for\ the\ Wikidata\ Query\ Service\ {[}@url:https://query.wikidata.org{]}\ \ It\ also\ allows\ users\ to\ feed\ an\ arbitrary\ list\ of\ articles\ and\ obtain\ a\ custom\ dashboard.\ \ Wikidata\ Bib\ obtains\ the\ html\ dashboards\ after\ feeding\ wbib\ the\ lists\ of\ articles\ read\ in\ total\ (}index.html\texttt{)\ or\ in\ pre-determined\ time\ spans\ (}last\_day.html\texttt{,}past\_week.html\texttt{and}past\_month.html` )

@@

The dashboard includes not only a basic list of read articles, but also statistics on most read authors and most read venues.
It also displays an interactive map of the instituions of articles read, permitting a glance on geographic biases in activities.
An example of queries is show in \ref{fig:dashboard}.
As the queries are rendered live, they evolve in quality with the growth of Wikidata.
Finally, the clean 5-star-open data format enables users to adapt the queries to include different aspects of Wikidata.
For example, table \ref{tbl:articles_read_hca} showcases 10 articles that (1) I have read in the past year and (2) were authored by a speaker of the 1st Human Cell Atlas Latin America Single Cell RNA-seqData Analysis Workshop {[}\protect\hyperlink{ref-1hag8XE6}{133}/{]}.
One practical application that the dashboard enables, thus, is to identify people in an event, institution or location that the user has read before, therefore catalysing the possibility of colaborations.
Anedoctally, this strategy was tested successfully at Biohackathon Europe 2021 {[}\protect\hyperlink{ref-kHL3NVxk}{134}/{]}, where I used the system both to identify possible collaborators and as a conversation starter.
\textbar workLabel \textbar authors \textbar{}
\textbar-------------------------------------------------------------------------------------------------\textbar--------------------------------------------------------------------------\textbar{}
\textbar A promoter-level mammalian expression atlas \textbar Jay W Shin \textbar{}
\textbar Single-cell RNA-seq reveals new types of human blood dendritic cells, monocytes, and progenitors.\textbar Muzlifah Haniffa \textbar{}
\textbar The Human Cell Atlas. \textbar Musa Mhlanga, Jay W Shin, Muzlifah Haniffa, Menna R Clatworthy, Dana Pe'er\textbar{}
\textbar The Human Cell Atlas: Technical approaches and challenges. \textbar Jay W Shin \textbar{}
\textbar Innate Immune Landscape in Early Lung Adenocarcinoma by Paired Single-Cell Analyses. \textbar Dana Pe'er \textbar{}
\textbar Single cell RNA sequencing of human liver reveals distinct intrahepatic macrophage populations \textbar Sonya A MacParland \textbar{}
\textbar Single-cell reconstruction of the early maternal--fetal interface in humans \textbar Muzlifah Haniffa \textbar{}
\textbar Distinct microbial and immune niches of the human colon \textbar Rasa Elmentaite, Menna R Clatworthy \textbar{}
\textbar A cell atlas of human thymic development defines T cell repertoire formation \textbar Muzlifah Haniffa, Menna R Clatworthy \textbar{}
\textbar Decoding human fetal liver haematopoiesis \textbar Muzlifah Haniffa \textbar{}
Table: Articles read by Tiago Lubiana before 8 December 2021 in which an author was a speaker at HCA Latin America
\{\#tbl:articles\_read\_hca\} \textbar{}
